External validation

### 2020-12-01 diagnostics/README.Rmd
m

---
title: "External validation"
author: "Erik Bulow"
date: '`r Sys.Date()`'
output: 
  github_document:
    toc: yes
    df_print: "kable"
bibliography: 
  C:\\Users\\erik_\\Documents\\library.bib
---

>>>>

---
title: "External validation"
author: "Erik Bulow"
date: '`r Sys.Date()`'
output: 
  github_document:
    toc: yes
    df_print: "kable"
bibliography: 
  # C:\\Users\\erik_\\Documents\\library.bib
  H:\\zotero_lib.bib
---

### 2020-12-01 diagnostics/external_validation_template.Rmd
m

---
title: "External validation"
author: "Erik Bulow"
date: '`r Sys.Date()`'
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: show
bibliography: 
  C:\\Users\\erik_\\Documents\\library.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE)

load("../cache/ext_val_required_data.RData")
load("../cache/tab_categorization.RData")
load("../cache/df.RData")
```

@@

```{r}
# First, set random seed for reproducability:
set.seed(123)

# USeful packages
pkgs <- c("tidyverse", "doParallel", "pROC", "rms", "givitiR")

# Install if not already installed
pkgsinst <- setdiff(pkgs, rownames(installed.packages()))
if (length(pkgsinst)) install.packages(pkgsinst)

@@

Load the exported models to validate (the R object previously sent, probably with another path). This is a nested data frame with one row for each prediction period (90 days and 2 years). Column `fit` contains `glm` object stripped from any personal patient data.

```{r}
# Load the exported model object
load("../cache/fit_brl_reduced_lean.RData")
fit_brl_reduced_lean
```

Let's inspect the model for 90 days just to get a sence of it:

```{r}
fit_brl_reduced_lean$fit[[1]]
```

We should now use this model with the `predict` function combined with new data from Denmark. So, how should this data look like?

@@

Those were the inclusions/exclusions from Sweden. Similar criteria should apply also for the external validation data set in to get comparable cohorts.

Ignore filtering of missing educational level, hospital type and fixation, however, since those variables are not needed in the model to validate.

@@

The outcome variables are simply called `outcome` in each model (hence, adjusted for each model). Those are logical/boolean (or 0/1-numeric) indicating whether the patient got PJI within 90 days (2 years) after THA (`TRUE`/1) or not (`FALSE`/0). Note that we had previously excluded all patients who died within two years. A competing risk model for survival analysis might be better but we were pragmatic in this case.

@@

The data to evaluate (in addition to the respective `outcome` variable) should look like this:

```{r}
# Object I use on my end, which includes patient data and can therefore not be shared
head(ext_val_required_data) 
```

thus with columns:

```{r}
names(ext_val_required_data)
```
Some of those are factor variables:
```{r}
ext_val_required_data %>% 
  select(is.factor) %>% 
  pivot_longer(everything(), values_ptypes = list(value = character())) %>% 
  distinct() %>% 
  arrange(name, value) %>% 
  group_by(name) %>% 
  mutate(name = replace(name, duplicated(name), ""))
```
- `P_Sex` and `P_ASA` should be self-explanatory
- `P_BMI` is a broader categorization based on BMI and the [WHO classification] (https://www.euro.who.int/en/health-topics/disease-prevention/nutrition/a-healthy-lifestyle/body-mass-index-bmi) where overweight = "pre-obesity"  
- `P_DiaGrp` is based on ICD-10 codes recorded in SHAR and grouped into broader categories as found in a separate CSV-file:

```{r}
readr::read_csv2("../data/P_DiaGrp.csv", trim_ws = TRUE)
```

@@

Variables prefixed by `c_`:

```{r}
nms     <- names(ext_val_required_data)
comorbs <- nms[startsWith(nms, "c_")]
comorbs
```
are logical/boolean indicators of comorbidities based on ICD-10/ATC codes from one year prior to THA, as recorded in our National Patient Register and medical prescription register. INdividual codes were grouped according to Charlson and Elixhauser as codified by table 2 in @Quan2005, and as RxRisk V according to table 1 in @Pratt2018. Those conditions were then further combined according to table 1 in the drafted manuscript:

```{r}
comorbs_nms <- 
  comorbs %>% 
  tolower() %>% 
  {gsub("^c_", "", .)} %>%
  {gsub("_", " ", .)}

tab_categorization %>% 
  filter(
    tolower(`Comorbidities by groups`) %in% comorbs_nms
  )
```

## Example validation for model of 90 days

We have the 90 day model from the shared R object;
```{r}
model <- fit_brl_reduced_lean$fit[[1]]
```

Let's assume we now have the `outcome` variable and a data frame `X` with the  predictors (I will use the Swedish data just as an example but this should of course be changed for the external validation).
```{r}
outcome <- df$outcome_infection_90d
X       <- ext_val_required_data
```

@@

```{r}
# Tibble with observed and predicted outcome
obspred <- 
  tibble(
    obs  = outcome, 
    pred = predict(model, X, type = "response")
  )

@@

# Estimate CI for AUC based on bootstrapping
# Use parallel processing to speed up the process
doParallel::registerDoParallel()
AUCci <- 
  pROC::ci.auc(
    ROC, 
    method          = "bootstrap", 
    boot.n          = 100,     # Remove this argument and use the default!
    boot.stratified = FALSE, 
    parallel        = TRUE
  )

# Check calibration. Note that devel should actually be "internal" for this example but I use
# "external", since that's what you will use for the external validation. 
calibration <- 
  givitiR::givitiCalibrationBelt(
    as.numeric(obspred$obs), 
    obspred$pred, 
    devel = "external"
  )
```

@@

A calibration belt plot might be illustrated as (this example might look strange due to the intentional internal/external miss-specification above):

```{r}
plot(calibration, xlim = c(0, 0.3), ylim = c(0, 0.3))
```

@@

Method 2 from table 1 in @Steyerberg2004.

```{r}
Z <- predict(model, X, type = "response")

# Refit the intercept using Z = a + Xb from above as offset
fit2 <- glm(outcome ~ 1, offset = Z)

# Same calibration and validation as above
obspred2     <- tibble(obs  = outcome, pred = predict(fit2, type = "response"))
ROC2         <- pROC::roc(obspred2, "obs", "pred", direction = "<")
AUCci2 <- 
  pROC::ci.auc(
    ROC2, 
    method = "bootstrap", 
    boot.n = 100,     # Remove this argument and use the default!
    boot.stratified = FALSE, 
    parallel = TRUE
  )

calibration2 <- 
  givitiR::givitiCalibrationBelt(
    as.numeric(obspred2$obs), 
    obspred2$pred, 
    devel = "external"
  )
```

@@

Method 3 from table 1 in @Steyerberg2004.

```{r}
fit3         <- glm(outcome ~ 1 + Z)
obspred3     <- tibble(obs  = outcome, pred = predict(fit3, type = "response"))
ROC3         <- pROC::roc(obspred3, "obs", "pred", direction = "<")

AUCci3 <- 
  pROC::ci.auc(
    ROC3, 
    method = "bootstrap", 
    boot.n = 100,     # Remove this argument and use the default!
    boot.stratified = FALSE, 
    parallel = TRUE
  )

calibration3 <- 
  givitiR::givitiCalibrationBelt(
    as.numeric(obspred3$obs), 
    obspred3$pred, 
    devel = "external"
  )
```

@@

roc_plot_coords <- 
  tibble(
    model = c("ias is", "re-calibrated intercept", "recalibrated"),
    data = 
      list( 
        tibble(
          specificities = ROC$specificities, 
          sensitivities = ROC$sensitivities
        ),
        tibble(
          specificities = ROC2$specificities, 
          sensitivities = ROC2$sensitivities
        ),
        tibble(
          specificities = ROC3$specificities, 
          sensitivities = ROC3$sensitivities
        )
      )
  )

@@

The text output from `AUCci`, `AUCci2` and `AUCci3` should be enough. Hence, the same character string that gets printed above (but now stored in an object).
```{r}
AUCci_print  <- capture.output(AUCci)
AUCci2_print <- capture.output(AUCci2)
AUCci3_print <- capture.output(AUCci3)
```

@@

Save objects above to file `export_90d.RData` (in the current working directory).

```{r}
save(
  roc_plot_coords,
  AUCci_print,
  AUCci2_print,
  AUCci3_print,
  calibration,
  calibration2,
  calibration3,
  file = "export_90d.RData"
)
```

# 2 years

Repeat for the 2-year model.

# Bibliography

>>>>

---
title: "External validation"
author: "Erik Bulow"
date: '`r Sys.Date()`'
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE)
```

@@

```{r}
pkgs <- c("tidyverse", "doParallel", "pROC", "rms", "givitiR")

# Install
pkgsinst <- setdiff(pkgs, rownames(installed.packages()))
if (length(pkgsinst)) install.packages(pkgsinst)

@@

Load the exported model to validate (the R object previsously sent, probably with another path).

```{r}
# Load the exported model object
load("../cache/fit_export.RData")
```

Set random seed for reproducability:

```{r}
set.seed(123)
```

@@

Those were the inclusions/exklusions from Sweden. It might not be necessary to filter out on BMI, hospital and education however. Those variables are not used in the model.

Additional filter to ages 35-99 years to match the Swedish cohort.

@@

The outcome variable is boolean (or numeric) indicating wether the patient died of any cause within 90 days after THA (`TRUE`/1) or not (`FALSE`/0). We did not have any censoring in Sweden. I guess we can simply drop cases where status is unknown?

@@

The data to evaluate should look like this:

```{r}
head(fit_export$data)
```

## Baseline variables

- **P_Gender**: Kvinna/Man = Female/Male
- **P_ASA**: level 1-3
- **P_Age**: 35 - 99

@@

Comorbidities are identified by Elixhauser and Charlson comorbidity during one year before surgery. This is done by ICD-10 codes in Sweden but other versions exist. If ICD-10-codes are used those might be identified by regular expressions in the table below. Those are based on codes like "C123". Hence no puntctuation (not "C12.3"). The regular expressions might be modified to allow possible other patterns. Regular expressoins below are just combined from differnet groups (separated by "|"). They could be rewritten for clarity. If you don't see any regular expressions, click the small arrow in the upper right corner of the table. The `grepl` function (base R) can be used to identify ICD-10-codes based on the regular expressions.

```{r, echo = FALSE}
load("../cache/categorization.RData")
comorbidities <- 
  categorization %>% 
  map_df(as.character) %>% 
  filter(new %in% names(fit_export$data)[grepl("c_", names(fit_export$data))]) %>% 
  separate(old, letters, sep = "\\|", fill = "right") %>% 
  pivot_longer(letters, values_drop_na = TRUE) %>% 
  mutate(value = trimws(value)) %>% 
  separate(value, c("index", "name"), extra = "merge") %>% 
  mutate(name = gsub("_", " ", name))

CCI <- 
  comorbidities %>% 
  filter(index == "CCI") %>% 
  left_join(coder::charlson_icd10, c(name = "group"))

ECI <- 
  comorbidities %>% 
  filter(index == "ECI") %>% 
  left_join(coder::elix_icd10, c(name = "group"))

comorb_defs <- 
  bind_rows(CCI, ECI) %>% 
  select(new, regex) %>% 
  mutate(regex = sprintf("(%s)", gsub("^", "", regex, fixed = TRUE))) %>% 
  group_by(new) %>% 
  summarise(regex = paste(regex, collapse = "|"))

comorb_defs
```

## Example data

Let's assume we now have some data (I will use the Swedish data just as an example).
```{r}
y <- fit_export$y
X <- fit_export$data
```

@@

```{r}
# Tibble with observed and predicted outcome
obspred <- 
  tibble(
    obs  = y, 
    pred = predict(fit_export, X, type = "response")
  )

@@

# Estimate CI for AUC based on bootstrapping
# Use parallel processing to speed up the process
doParallel::registerDoParallel()
AUCci <- 
  pROC::ci.auc(
    ROC, 
    method          = "bootstrap", 
    boot.stratified = FALSE, 
    parallel        = TRUE
  )

# Check calibration. Note that devel should actually be "internal" for this example but I use
# "external", since that's what you will use for the UK data. 
calibration <- 
  givitiR::givitiCalibrationBelt(
    obspred$obs, 
    obspred$pred, 
    devel = "external"
  )
```

@@

A calibration belt plot might be illustrated as:

```{r}
plot(calibration, xlim = c(0, 0.03), ylim = c(0, 0.06))
```

@@

Method 2 from table 1 in Steyerberg 2004.

```{r}
Z <- predict(fit_export, X, type = "response")

# Refit the intercept using Z = a + Xb from above as offset
fit2 <- glm(y ~ 1, offset = Z)

# Same calibration and validation as above
obspred2     <- tibble(obs  = y, pred = predict(fit2, type = "response"))
ROC2         <- pROC::roc(obspred2, "obs", "pred", direction = "<")
AUCci2       <- pROC::ci.auc(ROC2, method = "bootstrap", boot.stratified = FALSE, parallel = TRUE)
calibration2 <- givitiR::givitiCalibrationBelt(obspred2$obs, obspred2$pred, devel = "external")
```

@@

Method 3 from table 1 in Steyerberg 2004.

```{r}
fit3         <- glm(y ~ 1 + Z)
obspred3     <- tibble(obs  = y, pred = predict(fit3, type = "response"))
ROC3         <- pROC::roc(obspred3, "obs", "pred", direction = "<")
AUCci3       <- pROC::ci.auc(ROC3, method = "bootstrap", boot.stratified = FALSE, parallel = TRUE)
calibration3 <- givitiR::givitiCalibrationBelt(obspred3$obs, obspred3$pred, devel = "external")
```

@@

roc_plot_coords <- 
  data.frame(
    specificities = ROC$specificities, 
    sensitivities = ROC$sensitivities
  )

roc3_plot_coords <- 
  data.frame(
    specificities = ROC3$specificities, 
    sensitivities = ROC3$sensitivities
  )

@@

The text output from `AUCci` and `AUCci3` should be enough. Hence, the same character string that gets printed above (but now stored in an object).
```{r}
AUCci_print  <- capture.output(AUCci)
AUCci3_print <- capture.output(AUCci3)
```

@@

Save objects above to file `export.RData` (in the current working directory).

```{r}
save(
  roc_plot_coords,
  roc3_plot_coords,
  AUCci_print,
  AUCci3_print,
  file = "export.RData"
)
```

## Calibration plots

### Help function

This is a simple help function to make a clean calibration belt plot and save it as TIFF (in the curent working directory):

```{r}
makeplot <- function(x, file_name = deparse(substitute(x))) {
  tiff(
    paste0(file_name, ".tiff"), 
    1024, 1024, pointsize = 36, 
    compression = "lzw"
  )

  tcks <- seq(.0, .1, .01)

  plot(
    x,
    xlim             = c(0, .06),
    ylim             = c(0, .08),
    xlab             = "Predicted probabilities [%]",
    ylab             = "Observed  probabilities [%]",
    main             = NULL,
    table            = FALSE,
    polynomialString = FALSE,
    pvalueString     = FALSE,
    nString          = FALSE,
    mar              = c(5, 4, 0, 0) + 0.1,
    xaxt             = "n",
    yaxt             = "n"
  )
  abline(v = .03, lty = "dashed", col = "darkgreen", lwd = 3)
  axis(1, at = tcks, lab = sprintf("%.0f", tcks * 100), las = TRUE)
  axis(2, at = tcks, lab = sprintf("%.0f", tcks * 100), las = TRUE)

  dev.off()
}
```
### Make and save figures

```{r}
makeplot(calibration)
makeplot(calibration3)
```

## ROC plots

If it not possible to export data for ROC-plot, here is some code to make a figure and save it as TIFF (in the current working directory) instead:

roc_plot_coords %>% 
  ggplot(aes(1 - specificities, sensitivities)) +
  geom_path(size = 2) +
  geom_abline(intercept = 0, slope = 1, color = "grey", linetype = 2) +
  theme_minimal() +
  theme(
    legend.position = c(1, 0),
    legend.justification = c(1, 0),
    legend.title = element_blank()
  )

ggsave(
  "roc.tiff", 
  height = 10, 
  width = 10, 
  unit = "cm", 
  dpi = 900, 
  compression = "lzw"
)
```
